---
title: "03 - Simple linear regression"
output: html_notebook
---

```{r, echo = FALSE, message = FALSE}
library(openintro)
library(Stat2Data)
library(ggplot2)
library(dplyr)
library(psych)
data('galton')
data('possum')
data("ncbirths")
data('mammals')
data('bdims')
data('smoking')
mlbBat10 <- mlbbat10
Anscombe <- data.frame(x = c(anscombe$x1, anscombe$x2, anscombe$x3, anscombe$x4), y = c(anscombe$y1, anscombe$y2, anscombe$y3, anscombe$y4), set = rep(c(1, 2, 3, 4), each = 11))
noise <- data.frame(x = rnorm(1000), y = rnorm(1000), z = rep(1:20, 50))
galton <- read.csv('/home/cla/Documentos/Vitor/DataCamp/Data-Scientist-with-R/Datasets/galton.csv')
```

## **Visualization of Linear Models**

**1. Visualization of Linear Models**

Before we get into the mathematical specification for a regression model, let's build some intuition about what a regression line is.

**2. Possums**

In this scatter plot, we see the relationship between the total length of a possum and the corresponding length of its tail. Clearly there is not a perfect relationship here: the total length of the possum varies even for possums with the same tail length. But we still have some intuitive desire to describe the relationship with a line.

```{r}
ggplot(data = possum, aes(y = total_l, x = tail_l)) + 
  geom_point()
```

**3. Through the origin**

This line goes through the origin, that is, the point where both x and y are equal to zero and has a slope of 2 (point) 5 centimeters per centimeter. In some sense, it does go "through" the points, but doesn't capture the general trend as best we could imagine.

```{r}
ggplot(data = possum, aes(y = total_l, x = tail_l)) + 
  geom_point() + geom_abline(intercept = 0, slope = 2.5)
```

**4. Through the origin, better fit**

This line also goes through the origin, but has a gentler slope. It seems like a better fit since it cuts through the points in a more central way.

```{r}
ggplot(data = possum, aes(y = total_l, x = tail_l)) + 
  geom_point() + geom_abline(intercept = 0, slope = 2.3)
```

**5. Not through the origin**

But why should we force the line to go through the origin? Here is a line that has a y intercept of 40 centimeters and an even gentler slope of 1 (point) 3 centimeters per centimeter. It seems like an even better fit still. Do you think you could find an even better fit? In order to do so, you need some criteria for judging which line fits better. In particular, you need a numerical measurement of how good the fit of each possible line is. In regression, we use the least squares criterion to determine the best fit line.

```{r}
ggplot(data = possum, aes(y = total_l, x = tail_l)) + 
  geom_point() + geom_abline(intercept = 40, slope = 1.3)
```

**6. The "best" fit line**

Statisticians have proven that, apart from pathological examples, if we seek the line that tries to minimize the sum of the squared distances between the line and a set of data points, a unique line exists. That line is called the least squares regression line. We can add the line to our plot using the geom smooth function and specifying the method argument to be lm, which stands for linear model. Note that by default this will draw the regression line in blue with gray shading for the standard error associated with the line. That needn't concern us just yet,

```{r}
ggplot(data = possum, aes(y = total_l, x = tail_l)) + 
  geom_point() + geom_smooth(method = 'lm')
```

**7. Ignore standard errors**

so we can turn it off by setting the se argument to FALSE or to 0.

```{r}
ggplot(data = possum, aes(y = total_l, x = tail_l)) + 
  geom_point() + geom_smooth(method = 'lm', se = FALSE)
```

**8. Let's practice!**

You'll explore the best fit line on your own in these next exercises.

### **The "best fit" line**

The simple linear regression model for a numeric response as a function of a numeric explanatory variable can be visualized on the corresponding scatterplot by a straight line. This is a "best fit" line that cuts through the data in a way that minimizes the distance between the line and the data points.

We might consider linear regression to be a specific example of a larger class of *smooth* models. The `geom_smooth()` function allows you to draw such models over a scatterplot of the data itself. This technique is known as visualizing the model *in the data space*. The `method` argument to `geom_smooth()` allows you to specify what class of smooth model you want to see. Since we are exploring linear models, we'll set this argument to the value `"lm"`.

Note that `geom_smooth()` also takes an `se` argument that controls the standard error, which we will ignore for now.

**Instructions**

Create a scatterplot of body weight as a function of height for all individuals in the `bdims` dataset with a simple linear model plotted over the data.

```{r}
# Scatterplot with regression line
ggplot(data = bdims, aes(x = hgt, y = wgt)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = FALSE)
```

### **Uniqueness of least squares regression line**

The least squares criterion implies that the slope of the regression line is unique. In practice, the slope is computed by R. In this exercise, you will experiment with trying to find the optimal value for the regression slope for weight as a function of height in the `bdims` dataset via trial-and-error.

To help, we've built a custom function for you called `add_line()`, which takes a single argument: the proposed slope coefficient.

**Instructions**

The `bdims` dataset is available in your workspace. Experiment with different values (to the nearest integer) of the `my_slope` parameter until you find one that you think fits best.

```{r, echo = FALSE}
add_line <- function (my_slope) {

  bdims_summary <- bdims %>%
    summarize(N = n(), r = cor(hgt, wgt),
              mean_hgt = mean(hgt), mean_wgt = mean(wgt),
              sd_hgt = sd(hgt), sd_wgt = sd(wgt)) %>%
    mutate(true_slope = r * sd_wgt / sd_hgt, 
           true_intercept = mean_wgt - true_slope * mean_hgt)
  p <- ggplot(data = bdims, aes(x = hgt, y = wgt)) + 
    geom_point() + 
    geom_point(data = bdims_summary, 
               aes(x = mean_hgt, y = mean_wgt), 
               color = "red", size = 3)
  
  my_data <- bdims_summary %>%
    mutate(my_slope = my_slope, 
           my_intercept = mean_wgt - my_slope * mean_hgt)
  p + geom_abline(data = my_data, 
                  aes(intercept = my_intercept, slope = my_slope), color = "dodgerblue")
}
```

```{r}
# Estimate optimal value of my_slope
add_line(my_slope = 1.018)
```

## **Understanding Linear Models**

**1. Understanding Linear Models**

Models are ubiquitous in statistics.

**2. Generic statistical model**

In many cases, we assume that the value of our response variable is some function of our explanatory variable plus some random noise. The latter term is important and, in a philosophical sense, is the signal of statistical thinking. What we are saying here is that there is some mathematical function f, which can translate values of one variable into values of another, except that there is some randomness in the process. What often distinguishes statisticians from other quantitative researchers is the way that we try to model that random noise.

**3. Generic linear model**

For a linear regression model, we simply assume that f takes the form of a linear function. Thus, our model describes the value of the response variable in terms of an intercept and a slope.

**4. Regression model**

Mathematically, we usually call the intercept beta naught, and the slope beta one. The noise term is often denoted epsilon. In a regression model, we specify that the distribution of the noise is normal, with mean 0 and a fixed standard deviation. Again, understanding the specification of this noise term is crucial to thinking like a statistician. The part of the function aside from the noise term consists of a linear function that produces the fitted values.

**5. Fitted values**

These are usually denoted with Y hat. The difference between Y and Y hat, is that Y is the actual observed values of the response, while Y hat is the expected values of the response based on the model.

**6. Residuals**

The difference between those two quantities are called residuals. The order of the terms here matters, so that observations that exceed their expected value of the response variable will have positive residuals, while those that don't reach their expected value will have negative residuals. The residuals are the realization of the noise term. While epsilon and e play analogous roles in the regression equation, epsilon is an unknown, true quantity, while e is a known, estimate of that quantity. While we won't delve into the specifics of the least squares fitting procedure,

**7. Fitting procedure**

we will outline what it does. Given n observations of x,y pairs, the procedure finds estimates for the intercept and slope that minimize the sum of the squared residuals. The estimated intercept and slope are usually denoted beta naught hat and beta one hat, respectively. Note that here again, the hat indicates an estimated rather than a true quantity. Actually computing the fitted coefficients is usually a messy business for which the computer is much better suited, so we won't discuss that here.

**8. Least squares**

You should know that the least squares fitting procedure is a well-understood, relatively simple, deterministic process that can be computed efficiently. It will always return a unique solution, except in rare, unrealistic cases. The residuals are guaranteed to sum to 0, and the point, x bar, y bar, is guaranteed to lie on the regression line. Given your understanding of correlation, it should not surprise you to learn that the regression slope and the correlation coefficient are closely related. In fact, they are proportional to one another. You should also be aware that other criteria apart from least squares exist, but we won't talk about them in this course.

**9. Key concepts**

It's worth reviewing some key concepts about regression models. First, the fitted values, y hat, are the expected values given the corresponding value of x. It's not the case that x will always translate into y hat; it's just that y hat is our best guess for the true value of y given what we know about x. In the same way that the fitted intercept and slope are estimates of the true, unknown intercept and slope, the residuals are estimates of the true, unknown noise. Speaking of which, you may hear these noise terms referred to as errors. I tend to shy away from this terminology, since it suggests that something is wrong with your model. But there is not necessarily anything wrong, it's just that physical processes are subject to some random variation. All we are hoping to do is accurately capture that random noise.

**10. Let's practice!**

You'll put your understanding of regression to use in these next exercises.

### **Fitting a linear model "by hand"**

Recall the simple linear regression model:

$$Y = b_{0} + b_{1}.X$$
Two facts enable you to compute the slope $b_{1}$ and intercept $b_{0}$ of a simple linear regression model from some basic summary statistics.

First, the slope can be defined as:

 $$b_{1} = r_{X, Y}.\frac{s_{Y}}{s_{X}}$$

where $r_{X, Y}$ represents the correlation (`cor()`) of  and $X$ and $Y$ and $s_{X}$ and $s_{Y}$ represent the standard deviation (`sd()`) of $X$ and $Y$, respectively.

Second, the point ($\overline{x}$, $\overline{y}$) is always on the least squares regression line, where 
$\overline{x}$ and $\overline{y}$ denote the average of  and $X$ and $Y$, respectively.

The `bdims_summary` data frame contains all of the information you need to compute the slope and intercept of the least squares regression line for body weight ($Y$) as a function of height ($X$). You might need to do some algebra to solve for $b_{0}$!

**Instructions**

- Print the `bdims_summary` data frame.
- Use `mutate()` to add the `slope` and `intercept` to the `bdims_summary` data frame.

```{r, echo = FALSE}
bdims_summary <- data.frame(n = 507, r = 0.7173011, mean_hgt = 171.1438, sd_hgt = 9.407205, mean_wgt = 69.147553, sd_wgt = 13.34576)
```

```{r}
# Print bdims_summary
bdims_summary

# Add slope and intercept
bdims_summary %>%
  mutate(slope = r*(sd_wgt/sd_hgt), 
         intercept = mean_wgt - slope*mean_hgt)
```

## **Regression vs. regression to the mean**

**1. Regression vs. regression to the mean**

In the late 19th century, Sir Francis Galton developed much of the theory surrounding correlation and regression.

**2. Heredity**

One of the more enduring concepts is "regression to the mean." Here we note that "regression to the mean" is a distinct concept from "linear regression", which is the focus of this course. But "regression to the mean" is an important statistical concept in its own right, so we will explore it here so that you can appreciate the difference. As always, it's best to start with a question: do tall men tend to beget tall sons? Do tall women tend to have tall daughters? You might consider these questions in the context of professional basketball players. For example, do you suspect children of seven foot men to also be exceptionally tall? The answer turns out to be "yes, but not as tall as their fathers." This effect is what Galton deemed "regression to the mean".

**3. Galton's data**

In the plot, we see the relationship between the heights of a group of 465 men that Galton measured in the 1880s and their fathers. The height of the son is on the y axis, while the height of the corresponding father is on the x axis. The diagonal indicates sons who are as tall as their fathers. The slope of the regression line shown in blue is more gentle. This reflects the observation that while tall fathers are still likely to have tall sons, the sons are not likely to be as tall as their fathers. Their height has been "regressed" towards the mean height. It may be tempting to think that regression to the mean implies that men are getting shorter, but this is not true. Even in this sample, the mean height of the sons was the same to the nearest tenth of an inch to those of the fathers.

**4. Regression modeling**

This course is about simple linear regression modeling. Typically, regression models combine some explanatory variables into an estimate for a single, numerical response variable. We are pursuing least squares regression models, but there are any number of alternative regression model specifications. Additionally, there are models, like regression trees, that also estimate a single, numerical response, but do not use a linear framework as we do in this course.

**5. Let's practice!**

The next exercises will allow you to explore Galton's data on your own.

### **Regression to the mean**

*Regression to the mean* is a concept attributed to Sir Francis Galton. The basic idea is that extreme random observations will tend to be less extreme upon a second trial. This is simply due to chance alone. While "regression to the mean" and "linear regression" are not the same thing, we will examine them together in this exercise.

One way to see the effects of regression to the mean is to compare the heights of parents to their children's heights. While it is true that tall mothers and fathers tend to have tall children, those children tend to be less tall than their parents, relative to average. That is, fathers who are 3 inches taller than the average father tend to have children who may be taller than average, but by less than 3 inches.

The `Galton_men` and `Galton_women` datasets contain data originally collected by Galton himself in the 1880s on the heights of men and women, respectively, along with their parents' heights.

Compare the slope of the regression line to the slope of the diagonal line. What does this tell you?

**Instructions**

- Create a scatterplot of the height of men as a function of their father's height. Add the simple linear regression line and a diagonal line (with slope equal to 1 and intercept equal to 0) to the plot.
- Create a scatterplot of the height of women as a function of their mother's height. Add the simple linear regression line and a diagonal line to the plot.

```{r, echo = FALSE}
colnames(galton) <- c('family', 'father', 'mother', 'sex', 'height', 'nkids')
Galton_men <- galton %>% 
  filter(sex == ' M')
Galton_women <- galton %>% 
  filter(sex == ' F')
```

```{r}
# Height of children vs. height of father
ggplot(data = Galton_men, aes(x = father, y = height)) +
  geom_point() + 
  geom_abline(slope = 1, intercept = 0) + 
  geom_smooth(method = 'lm', se = FALSE)

# Height of children vs. height of mother
ggplot(data = Galton_women, aes(x = mother, y = height)) +
  geom_point() + 
  geom_abline(slope = 1, intercept = 0) + 
  geom_smooth(method = 'lm', se = FALSE)
```

Because the slope of the regression line is smaller than 1 (the slope of the diagonal line) for both males and females, we can verify Sir Francis Galton's regression to the mean concept!